---
title: "R Programming"
author: "Dr. Xiang Ji @ Tulane University"
date: "Sep 27, 2021"
output:
  html_document:
    toc: true
    toc_depth: 4  
subtitle: MATH-7360 Data Analysis
csl: ../apa.csl
---


```{r setup, include=FALSE}
rm(list = ls()) # clean-up workspace
knitr::opts_chunk$set(fig.align = 'center', cache = FALSE)
sessionInfo()
```

## Annoucement

- **One-page** project description due this Wednesday

- HW1 due this Friday.  Extension?

- Questions regarding HW1 problem 3 from Viraj (everyone should thank him)

  1. For creating the variable u, should we create it as a vector (using the c function) or as a matrix <br>
  _ans_: `u` is a column vector.
  2. For cases where we are multiplying a 6 by 6 matrix with a 1 by 1 or a 1 by 6 matrix, could we use the as.vector or the c() function on the 1 by 1 or the 1 by 6 matrix to compute the matrix-vector product? <br>
  _ans_: you can multiple a scalar to a matrix.  The result is an element-wise multiplication of the scalar with every entry in the matrix.  You might want to try as.numeric(scalar) * matrix.
  3.  Is problem 3e asking for the maximum of the sum of all the values in matrix U? <br>
  _ans_: It's asking for the maximum of row sums. (maximum over row index i, sum over column index j)
  4. Lastly for products where we are multiplying matrices, we should be using the matrix multiplication function (%*%) and not the element wise multiplication function, is that correct?  <br>
  _ans_: Yes.  When performing matrix-matrix / matrix-vector multiplications, use matrix operations with %*%.
  

## Acknowledgement

Dr. Hua Zhou's [slides](https://ucla-biostat203b-2020winter.github.io/slides/14-advr/advr1.html)

## _Advanced R_

To gain a deep understanding of how R works, the book [Advanced R](https://adv-r.hadley.nz) by Hadley Wickham is a must read. Read **now** to save numerous hours you might waste in future.

We cover select topics on coding style, benchmarking, profiling, debugging, parallel computing, byte code compiling, Rcpp, and package development.

## Style

- [Google's R style](https://google.github.io/styleguide/Rguide.xml).

- [Hadley Wickham's tidyverse style guide](https://style.tidyverse.org).

- Most important is to be consistent in your code style.

## Benchmark

Sources:  

- _Advanced R_: <http://adv-r.had.co.nz/Performance.html>, <https://adv-r.hadley.nz/perf-measure.html#microbenchmarking>   

- Blog: <http://www.alexejgossmann.com/benchmarking_r/>   

In order to identify performance issue, we need to measure runtime accurately.

### `system.time`

```{r}
set.seed(280)
x <- runif(1e6)

system.time({sqrt(x)})
system.time({x ^ 0.5})
system.time({exp(log(x) / 2)})
```

From [William Dunlap](http://r.789695.n4.nabble.com/Meaning-of-proc-time-td2303263.html#a2306691):

> "User CPU time" gives the CPU time spent by the current 
process (i.e., the current R session) and "system CPU time" 
gives the CPU time spent by the kernel (the operating 
system) on behalf of the current process.  The operating 
system is used for things like opening files, doing 
input or output, starting other processes, and looking at 
the system clock:  operations that involve resources that many 
processes must share.  Different operating systems will 
have different things done by the operating system.

### `microbenchmark`

```{r}
library("microbenchmark")
library("ggplot2")

mbm <- microbenchmark(
  sqrt(x),
  x ^ 0.5,
  exp(log(x) / 2),
  times = 100
)
mbm
```
Results from `microbenchmark` can be nicely plotted in base R or ggplot2.
```{r}
boxplot(mbm)
autoplot(mbm)
```

### `bench`

The `bench` package is another tool for microbenchmarking. The output is a tibble:
```{r}
library(bench)

(lb <- bench::mark(
  sqrt(x),
  x ^ 0.5,
  exp(log(x) / 2)
))
```
To visualize the result
```{r}
plot(lb)
```

It is colored according to GC (garbage collection) levels.

## Profiling

> Premature optimization is the root of all evil (or at least most of it) in programming.  
> -Don Knuth

Sources:   

- <https://adv-r.hadley.nz/perf-measure.html#profiling>  

- <https://rstudio.github.io/profvis/>  

- <https://support.rstudio.com/hc/en-us/articles/218221837-Profiling-with-RStudio>

### First example

```{r}
library(profvis)

profvis({
  data(diamonds, package = "ggplot2")

  plot(price ~ carat, data = diamonds)
  m <- lm(price ~ carat, data = diamonds)
  abline(m, col = "red")
})
```

### Example: profiling time

First generate test data:
```{r}
times <- 4e5
cols <- 150
data <- as.data.frame(x = matrix(rnorm(times * cols, mean = 5), ncol = cols))
data <- cbind(id = paste0("g", seq_len(times)), data)
head(data)
```
Original code for centering columns of a dataframe:
```{r}
profvis({
  # Store in another variable for this run
  data1 <- data
  
  # Get column means
  means <- apply(data1[, names(data1) != "id"], 2, mean)
  
  # Subtract mean from each column
  for (i in seq_along(means)) {
    data1[, names(data1) != "id"][, i] <-
      data1[, names(data1) != "id"][, i] - means[i]
  }
})
```

Profile `apply` vs `colMeans` vs `lapply` vs `vapply`:
```{r}
profvis({
  data1 <- data
  # Four different ways of getting column means
  means <- apply(data1[, names(data1) != "id"], 2, mean)
  means <- colMeans(data1[, names(data1) != "id"])
  means <- lapply(data1[, names(data1) != "id"], mean)
  means <- vapply(data1[, names(data1) != "id"], mean, numeric(1))
})
```

We decide to use `vapply`:
```{r}
profvis({
  data1 <- data
  means <- vapply(data1[, names(data1) != "id"], mean, numeric(1))

  for (i in seq_along(means)) {
    data1[, names(data1) != "id"][, i] <- data1[, names(data1) != "id"][, i] - means[i]
  }
})
```

Calculate mean and center in one pass:
```{r}
profvis({
 data1 <- data
 
 # Given a column, normalize values and return them
 col_norm <- function(col) {
   col - mean(col)
 }
 
 # Apply the normalizer function over all columns except id
 data1[, names(data1) != "id"] <-
   lapply(data1[, names(data1) != "id"], col_norm)
})
```

### Example: profiling memory

Original code for cumulative sums:
```{r}
data <- data.frame(value = runif(1e5))
profvis({
  data$sum[1] <- data$value[1]
  for (i in seq(2, nrow(data))) {
    data$sum[i] <- data$sum[i-1] + data$value[i]
  }
})
```

Write a function to avoid expensive indexing by `$`:
```{r}
profvis({
  csum <- function(x) {
    if (length(x) < 2) return(x)

    sum <- x[1]
    for (i in seq(2, length(x))) {
      sum[i] <- sum[i-1] + x[i]
    }
    sum
  }
  data$sum <- csum(data$value)
})
```

Pre-allocate vector:
```{r}
data <- data.frame(value = runif(1e6))
profvis({
  csum2 <- function(x) {
    if (length(x) < 2) return(x)

    sum <- numeric(length(x))  # Preallocate
    sum[1] <- x[1]
    for (i in seq(2, length(x))) {
      sum[i] <- sum[i-1] + x[i]
    }
    sum
  }
  data$sum <- csum2(data$value)
})
```

### Advice

Modularize big projects into small functions. Profile functions as early and as frequently as possible.

## Debugging

Learning sources:  

- Video: <https://vimeo.com/99375765>  

- _Advanced R_: <https://adv-r.hadley.nz/debugging.html>   

- breakpoint

- step in/through function

- `browser()`

- `traceback()`

- `options(error = browser)`, `options(error = NULL)`, `Debug` -> `On Error` -> `Break in Code`