---
title: "Lab 08 web scraping"
author: "Dr. Xiang Ji @ Tulane University"
date: "Oct 9, 2020"
output:
  html_document:
    toc: true
    toc_depth: 4  
subtitle: MATH-7360 Data Analysis
csl: ../apa.csl
---


### HTML scraping

We will scrape NFL data in this lab.

We will follow [this tutorial](https://towardsdatascience.com/scraping-nfl-stats-to-compare-quarterback-efficiencies-4989642e02fe) on scraping NFL stats (although the tutorial uses python).

We want to scrape from the webpage of [https://www.pro-football-reference.com/years/2019/passing.htm](https://www.pro-football-reference.com/years/2019/passing.htm)

Now open html with rvest

```{r}
rm(list = ls()) # clean-up workspace
library(tidyverse)
library(rvest)
url <- "https://www.pro-football-reference.com/years/2019/passing.htm"
(webpage <- read_html(url))
```

### Parse the table

Inspect the table elements and try to find the patterns you can use to make the CSS selector for fetching the rows of the table


The [CSS selector reference](https://www.w3schools.com/cssref/css_selectors.asp) could be helpful here.

### Get column names

Find the CSS selector for the title row of the table.  Is it an element of `thead`?

Now get all the column names passing elements inside the `thead` element.

```{r}
(col_names_html <- html_nodes(webpage, "thead"))
col_names <- html_text(col_names_html) %>%
  str_replace_all(c("\n" = "", "\\s+" = " ", "\"" = "")) %>%
  str_split(" ") %>%
  unlist() %>%
  print()

# I will just manually remove the last empty element
(col_names <- col_names[-length(col_names)])
```

### get each row of the table

Now find the CSS selector for each row of the table.

For example, one possible rule your instructor found is to look for all the \<tr\> tags inside the \<tbody\> element.


```{r}
# prepare patterns first
data_stat_names <- col_names %>%
  str_to_lower()
data_stat_names[c(1, 3, 8)] = c("ranker", "team", "qb_rec")
data_stat_names[9:16] <- str_c("pass", str_replace(data_stat_names[9:16], "%", "_perc"), sep = "_")
data_stat_names[17:23] <- str_c("pass", c("first_down", "long", "yds_per_att", "adj_yds_per_att", "yds_per_cmp", "yds_per_g", "rating"), sep = "_")
data_stat_names[25:29] <- str_c("pass", c("sacked", "sacked_yds", "net_yds_per_att", "adj_net_yds_per_att", "sacked_perc"), sep = "_")
data_stat_names[30] <- "comebacks"
print(data_stat_names)


# Now use the CSS selectors to scrape data and store them in a matrix
scraped_data <- NULL
for (stat_name in data_stat_names) {
  col_pattern <- str_c('[data-stat=', stat_name, ']', sep = "")
  (col_data_html <- html_nodes(webpage, str_c("tbody tr ", col_pattern, sep = "")))
  col_data <- html_text(col_data_html)
  scraped_data <- cbind(scraped_data, col_data)
}
colnames(scraped_data) <- col_names

scraped_data <- as_tibble(scraped_data) %>%
  filter(Rk != "Rk")
scraped_data

write_csv(scraped_data, "./NFL_passing_2019.csv")
```

