---
title: "Lab 04 ggplot2"
author: "Dr. Xiang Ji @ Tulane University"
date: "Oct 8, 2021"
output:
  html_document:
    toc: true
    toc_depth: 4  
subtitle: MATH-7360 Data Analysis
csl: ../apa.csl
---

```{r}
rm(list = ls()) # clean-up workspace
library("tidyverse")
```

## Some exercise on ggplot2

### Aesthetic mappings

Q1. What's gone wrong with this code?

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, colour = "blue"))
```

------------

Q2. Which variables in `mpg` are categorical? Which variables are continuous?
(Hint: type `?mpg` to read the documentation for the dataset).
How can you see this information when you run `mpg`?

------------

Q3. Map a continuous variable to `color`, `size`, and `shape`.
How do these aesthetics behave differently for categorical vs. continuous variables?

------------

Q4. What happens if you map the same variable to multiple aesthetics?

------------

Q5. What happens if you map an aesthetic to something other than a variable name, like `aes(colour = displ < 5)`? Note, you’ll also need to specify `x` and `y`.

------------

### Geometric objects

Q6. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

------------

Q7. Will these two graphs look different? Why/why not?

```{r, eval=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

> We will continue with ggplot exercise in next lab.
