---
title: "Lab 02 R Basics"
author: "Dr. Xiang Ji @ Tulane University"
date: "Aug 28, 2020"
output:
  html_document:
    toc: true
    toc_depth: 4  
subtitle: MATH-7360 Data Analysis
csl: ../apa.csl
---


> **Question 1** What type do the above vectors hold?

```{r}
a <- c(1L, "This is a character")
str(a)

b <- c(TRUE, "Hello World")
str(b)

c <- c(FALSE, 2)  # what is wrong here?
str(c)
```

 - In R, integers are specified by the suffix `L` (e.g. `1L` is an integer 1).  All other numbers are of class `numeric`.
   **Note** the function `is.integer` does not test whether a given variable has an integer value, but whether it belongs to the class integer.
  
 - So that integer or logical elements are coerced into characters to make `a` and `b` to be atomic vectors of characters.
 
 - `2` here generates a numeric (double) instead of integer (without the suffix `L`).  Logical element is coerced to numeric and `c` is an atomic vector of numeric.

## Explicit coercion

```{r}
# using the same objects a, b, c from the above question
a.logical <- as.logical(a)
a.integer <- as.integer(a)
a.numeric <- as.numeric(a)

b.logical <- as.logical(b)
b.integer <- as.integer(b)
b.numeric <- as.numeric(b)

c.logical <- as.logical(c)
c.integer <- as.integer(c)
c.numeric <- as.numeric(c)
c.character <- as.character(c)


d <- -5:5
d.logical <- as.logical(d)
```

> **Question 2** What do you get after the coercions?  Does any one suprise you?  Can you figure out why?

Vectors `a`, `b`, and `c` already went through coercions before we explicitly coerce them.
Therefore, when we explicitly coerce them, `a` and `b` are character vectors and `c` is numeric vector.

Some surprises:

 - Although `as.logical(1L) = TRUE`.  `as.logical("1") = FALSE`.
 
 - Only `0L` and `0` are coerced into `FALSE` by `as.logical()`.  All others are coerced to `TRUE`.

```{r}
a
str(a.logical)
str(a.integer)
str(a.numeric)

b
str(b.logical)
str(b.integer)
str(b.numeric)

c
str(c.logical)
str(c.integer)
str(c.numeric)

```


> **Question 3** Explain what you found.

First let's create a vector $\mathbf{v} = (969, 971, 972, \dots, 1022, 1023)$ of 54 elements

```{r}
# finish the code below
v <- c(969, 971:1023)
# or
v <- 969:1023
v <- v[-2]

length(v)
```

Then, let's compute the sum $\sum_{i=1}^{54}2^{v_i}$.

```{r}
# finish the code below
v.power.sum <- sum(2^v)
v.power.sum
```

How about only sum over 53 elements $\sum_{i=2}^{54}2^{v_i}$  (note that the sum starts from $v_2$).

```{r}
# finish the code below
v.power.sum.53 <- sum(2^v[-1])
v.power.sum.53
```

Now let's try putting the first element back

```{r}
v.power.sum.second <- 2^v[1] + sum(2^v[-1])
v.power.sum.second
```

Reference - this should be the bigget number
```{r}
biggest <- .Machine$double.xmax
biggest
```

Findings:

 - The sum $\sum_{i=1}^{54}2^{v_i}$ creates an **overflow** error for double
 
 - $\sum_{i=2}^{54}2^{v_i}$ does not create overflow error
 
 - `2^v[1] + sum(2^v[-1])` actually equals `sum(2^v[-1])`.  This is because adding two numbers of dramatically different magnitudes, we get `a + b = b`!
 It is referred to as Catastrophic cancellation caused by the finite-precision arithmetic.
 You can read more [here](https://en.wikipedia.org/wiki/Loss_of_significance).
